Lenses with a Difference

### 2018-06-22 papers/DifferenceLenses/main.tex
Added technical material

\newcommand{\kw}[1]{\ensuremath{\mathsf{#1}}\xspace}
\newcommand{\get}{\ensuremath{\kw{get}}\xspace}
\newcommand{\pput}{\ensuremath{\kw{put}}\xspace}
\newcommand{\create}{\ensuremath{\kw{create}}\xspace}
\newcommand{\res}{\ensuremath{\kw{res}}\xspace}
\newcommand{\ccopy}{\ensuremath{\kw{copy}}\xspace}
\newcommand{\const}{\ensuremath{\kw{const}}\xspace}
\newcommand{\default}{\ensuremath{\kw{default}}\xspace}
\newcommand{\length}{\ensuremath{\kw{length}}\xspace}
\newcommand{\match}{\ensuremath{\kw{match}}\xspace}

\DeclareMathOperator*{\argmin}{arg\,min} % thin space, limits underneath in
\DeclareMathOperator*{\argmax}{argmax} % no space, limits underneath in

@@

For example, assume that a source consists of a list of pairs of names, and that the target consists of a list of single names. Assume that $\ell$ is a bx which in the forward direction (i.e. the \get direction) simply discards the first of the two names:
$$\ell.\get \; [``Ada \; Lovelace", \; ``Alan \; Turing"] = [``Lovelace", \; ``Turing"]$$
If the order of the names in the target is reversed then a user might expect that in the backward direction (i.e. the \pput direction), the lens would simply restore the original names:
$$\ell.\pput \; [``Ada \; Lovelace", \; ``Alan \; Turing"] \; [``Turing", \; ``Lovelace"] = [``Ada \; Lovelace", \; ``Alan \; Turing"]$$
However, $\ell$ may not perform this alignment so that instead, the following result occurs:
$$\ell.\pput \; [``Ada \; Lovelace", \; ``Alan \; Turing"] \; [``Turing", \; ``Lovelace"] = [``Ada \; Turing", \; ``Alan \; Lovelace"]$$
Consider another example where both a source and a target consists of a list where every element on the list is a name and a year. Assume that $\ell$ is a bx which in the \get direction returns all elements of the source list that have year greater than or equal to 1900:
$$\ell.\get \; [``Ada \; Lovelace, \; 1815", \; ``Alan \; Turing, \; 1912"] = [``Alan \; Turing, \; 1912"]$$
Suppose that the entry $``Ada \; Lovelace, \; 1915"$ is then added to the target list. Should $\ell.\pput$ replace the old entry containing the name $``Ada \; Lovelace"$,
\begin{align*}
&\ell.\pput \; [``Ada \; Lovelace, \; 1815", \; ``Alan \; Turing, \; 1912"] \; [``Alan \; Turing, \; 1912", \; ``Ada \; Lovelace, \; 1915"] \\
&= [``Ada \; Lovelace, \; 1915", \; ``Alan \; Turing, \; 1912"]
\end{align*}
or should the \pput application retain both versions of $``Ada \; Lovelace"$?
\begin{align*}
&\ell.\pput \; [``Ada \; Lovelace, \; 1815", \; ``Alan \; Turing, \; 1912"] \; [``Alan \; Turing, \; 1912", \; ``Ada \; Lovelace, \; 1915"] \\
&= [``Ada \; Lovelace, \; 1815", \; ``Alan \; Turing, \; 1912", \; ``Ada \; Lovelace, \; 1915"]
\end{align*}
In both of these examples, there are multiple ways to restore consistency---that is, there are multiple ways to define $\ell.\pput \; s \; v$ in such a way that $\ell.\get \; (\ell.\pput \; s \; v) = v$---and the results from the different \pput applications may be vastly different from what a user expects.

@@

We take the approach that a bidirectional programming language should have explicit guarantees on how programs written in that language propagate updates; ideally these guarantees should be strictly stronger than hippocraticness. This way a programmer who derives bx transformations in this language has a clearer understanding of the behaviour of the programs that she writes.

In this paper we will show how this can be done, by describing a formalization of bx transformations that explicitly accounts for update propagation, and presenting a series of examples showing how this principle can be applied to existing bx programming languages. More specifically, the major technical contributions that we will make in this paper are the following:
\begin{enumerate}
\item
We present a formalization for bx transformations that requires a bx to provide guarantees on how it propagates updates. 
\item
We describe a specific bx language which we call {\em difference lenses}; this language is based on the {\em basic lenses} defined in the bx programming language {\em Boomerang}\cite{foster2009bidirectional}, but each difference lens comes with explicit guarantees on update propagation. As a result of this work, we will describe non-trivial update propagation properties that Boomerang's basic lenses provide.
  \item
  We show how to represent matching lenses and  relational lenses \cite{bohannon2006relational} as difference lenses, and hence describe non-trivial guarantees that these lens languages provide as well.
\end{enumerate}
\iffalse
 we describe a compositional language of lenses called {\em difference lenses}, where each difference $\ell$ satisfies the following equation,
\begin{equation}\label{difflenslaw}
\delta_S(s, \ell.\pput \; s \; v) = \delta_V(\ell.\get \; s, v)
\end{equation}
for appropriate real-valued $\delta_S : S \times S \longrightarrow \mathbb{R}^{\geq 0}$ and $\delta_S : S \times S \longrightarrow \mathbb{R}^{\geq 0}$.

\cref{difflenslaw}, which already appears in Gibbons et. al. \cite{gibbons2017principles} as the principle of {\em Weak Least Surprise}, is simple and intuitive but presents a challenge when trying to develop a compositional lens langauge that satisfies it. For example, assume that a source consists of a name-and-date pairs while a view consists of a name. Assume that a lens $\ell$ simply discards dates in the \get direction:
\begin{lstlisting}[mathescape=true]
$\ell.\get$ "Ada Lovelace, 12-10-1815" = "Ada Lovelace"
\end{lstlisting}
Consider the iteration lens $\ell^*$, where the source is now a list of name-and-date pairs while a view consists of a list of names. Whenever the length of a view list is larger than the length of a source list, then a \pput application of $\ell^*$ must initialize a source from a view from scratch:
\begin{lstlisting}[mathescape=true]
$\ell.\pput$ [] ["Ada Lovelace"] = ["Ada Lovelace, 00-00-0000"]
\end{lstlisting}
Now if we are to define $\delta_S^*$ and $\delta_V^*$ inductively on source and view lists respectively in such a way that \cref{difflenslaw} holds of $\ell^*$, then the following equation would have to hold:
\begin{lstlisting}[mathescape=true]
$\delta_S^*$(["Ada Lovelace, 00-00-0000"], []) = $\delta_V^*$(["Ada Lovelace"], [])
\end{lstlisting}
In a case such as this when a source is {\em created} from a view as opposed to arising from a view being \pput into an existing source, the source delta will be defined so as to defer to the view delta by mapping the created source back into the view; in particular, for the lens $\ell^*$ in the example, then we have
\begin{lstlisting}[mathescape=true]
$\delta_S^*$(["Ada Lovelace, 00-00-0000"], [])
= $\delta_V^*$([$\ell.\get$ "Ada Lovelace, 00-00-0000"], [])
= $\delta_V^*$(["Ada Lovelace"], [])
\end{lstlisting}

By now there exist many programming languages that enable one to define {\em
bidirectional transformations}, programs that can be run in two
directions. Each of these languages provides guarantees on how the two
transformations interact e.g. that they are bijective, that a round-trip on
one or both directions does not cause non-trivial updates etc. Many of these
guarantees have to do with restoring or preserving consistency. To formalize
this statement, we describe a widely used formulation of bidirectional
transformations due to Stevens \cite{stevens2010bidirectional} but originally
articulated in a more general form by Meertens \cite{meertens1998designing}:
\begin{definition}
A bidirectional transformation (bx) $R: M \Leftrightarrow N$ comprises of a
consistency relation $R \subseteq M \times N$ and restorers $\overrightarrow{R}
: M \times N \rightarrow N$ and $\overleftarrow{R} : M \times N \longrightarrow
N$.
\end{definition}
\begin{definition}
A bx $R : M \Leftrightarrow N$ is {\em correct} if the consistency restorers do
restore consistency: that is, for all $m \in M, n \in N$, we have $R(m,
\overrightarrow{R}(m, n))$, and dually. $R$ is {\em hippocratic} if the
consistency restorers make no change to an already-consistent pair of models:
that is, for all $m \in M$, $n \in N$, if $R(m, n)$ then $\overrightarrow{R}(m,
n) = n$ and dually. $R$ is {\em well behaved} if it is correct and hippocratic.
\end{definition}

Many bidirectional programming languages thus enable one to define
well behaved bx's: consistency restoration is guaranteed by correctness of the
bx while consistency preservation is guaranteed by the hippocraticness of the
bx. This paper is primarily concerned with well behaved bx's where
$\overrightarrow{R}$, the \get function of the bx, ignores it's $N$-argument and
where the consistency relation $R$ is defined by $R(m, \overrightarrow{R}(m))$
for all $m \in M$--- these transformations are the well known and widely
understood {\em lenses} of Foster et al \cite{foster2007combinators}--- though
the ideas we develop could be used in other bidirectional settings as well.

When the consistency relation of a bx $R$ is very fine, i.e. when each $m \in M$
is consistent with only a few $n \in N$ and vice versa, then a well behaved lens
gives fairly strong guarantees on $\overleftarrow{R}$ and $\overrightarrow{R}$.
However, there are many cases that arise in practice where there are many
different ways to restore consistency; when this happens, well-behavedness may
be too weak to rule out ``bad'' behaviour or to clarify the behaviour of a bx in
subtle cases. Various solutions have been proposed to address this issue, for
example {\em matching lenses} that deal with the issue of alignment
\cite{barbosa2010matching}, {\em edit lenses} \cite{hofmann2012edit} and {\em
delta lenses} \cite{diskin2011asymmetric,diskin2011state,pacheco2012delta}
that explicitly propagate edits through the lens, and {\em least change
lenses} that require that the source value resulting from a \pput application
is not only acceptable, but also one of closest to the original $s$ among the
sources that share the same view $v$ \cite{macedo2013composing}.

This paper takes a different approach to specifying update propagation by
identifying a condition called {\em $\partial D$ preservation} which we describe
below:
\begin{definition}
Let $R : M \Leftrightarrow N$ be a bidirectional transformation and $\partial D
\subseteq (M \times M) \times (N \times N)$. Then $R$ {\em preserves $\partial
D$} if for all $m \in M$, $n, n' \in N$ , if $R(m, n)$ then $\partial D((m,
\overleftarrow{R}(m, n')), (\overrightarrow{R}(m, n), n'))$. $R$ {\em
strongly preserves $\partial D$} if the dual condition also holds.
\end{definition}
\noindent The idea behind $\partial D$ preservation is that if $m$ and $n$ are
already consistent and $\overrightarrow{R}(m,n)$ is updated to produce $n'$,
then the difference between $\overrightarrow{R}(m,n)$ and $n'$ is related
to the difference between $m$ and $\overleftarrow{R}(m,n')$. $R$
strongly preserves $\partial D$ if it preserves $\partial D$ in both directions.

For lenses, $\partial D$ preservation says that difference between $\ell.\get \;
s$ and $v$ is related to the difference between $s$ and $\ell.\pput \; s \;
v$. One way to formalize this idea is to model the difference between two sources $s, s'
\in S$ and two views $v, v' \in V$ as functions $\delta_S : S \times S
\longrightarrow \partial S$ and $\delta_V : V \times V \longrightarrow \partial
V$ for some sets $\partial S$ and $\partial V$, and to capture the
relationship between $\delta_S$ and $\delta_V$ by a lens. In
particular, given a lens $\partial(S, V) : \partial S \Leftrightarrow \partial
V$ we will ask if the following equation holds of $\ell$:
\begin{equation}\label{difflenslaw}
\partial(S, V).\get \; (\delta_S(s, \ell.\pput \; s \; v)) = \delta_V(\ell.\get
\; s, v)
\end{equation}
If $\ell : S \Leftrightarrow V$ satisfies \cref{difflenslaw} we say that 
$\ell$ is a {\em difference  lens} with respect to $(\delta_S, 
\partial(S, V), \delta_V)$, which we denote with the typing relation 
$\ell : S \xLeftrightarrow{\delta_S, \partial(S, V), \delta_V} V$.

For example, consider two versions of an iteration lens $\ell_1 : S^*
\Leftrightarrow V^*$ and $\ell_2 : S^* \Leftrightarrow V^*$ defined inductively
from a lens $\ell : S \Leftrightarrow V$. Both of these lens have the same get
behaviour: given a list $s_1, \ldots, s_n$ of sources, they both produce
$\ell.\get \; s_1, \ldots, \ell.\get \; s_n$. The \pput behaviour of the two
lenses is different though: $\ell_1$ simply aligns views to sources
positionally, while $\ell_2$ computes the optimal alignment on the view side and
uses this alignment to put views back into sources.

Let \lstinline|name = [A-Z] . [a-z]*|, let the source $S$ equal 
\lstinline|name . " " . name| and the view $V$ equal \lstinline|name|.
Let $\ell : S \Leftrightarrow V$ be the lens
\lstinline|del name . del " " . copy name| which given a pair of names produces
the second name in the \get direction, and restores the first name in the \pput
direction:
$$\ell.\get \; ``Ada \; Turing" = ``Turing" \text{ and }\; \ell.\pput \; ``Ada
\; Turing" \; ``Lovelace" = ``Ada \; Lovelace"$$
Define the
function $\delta_V : V \times V \longrightarrow \mathbb{N}^{\geq 0}$ by 
$$\delta_V(n, n') = 
\begin{cases}
0 & \text{if }n=n'\\
1 & \text{otherwise}
\end{cases}$$
Define $\delta_S : S \times S \longrightarrow \mathbb{N}^{\geq 0}$ by
$$\delta_S(m \cdot `` \quad " \cdot n, m' \cdot `` \quad " \cdot n') =
\begin{cases}
0 & \text{if } m = m', n = n'\\
1 & \text{if } m = m', n \neq n'\\
1 & \text{if } m \neq m', n = n'\\
2 & \text{if } m \neq m', n \neq n'
\end{cases}$$
Let the lens $\partial(S, V) : \mathbb{N}^{\geq 0} \Leftrightarrow
\mathbb{N}^{\geq 0}$ be given by $\partial(S, V) = \ccopy(\mathbb{N}^{\geq
0})$. Then $\ell : S \xLeftrightarrow{\delta_S, \partial(S, V), \delta_V} V$.

Now let $xs, ys$ be two lists of single names, with $length(xs) \geq length(ys)$
Define $\delta_V^* : V^* \times V^* \longrightarrow \mathbb{Z}^* \times V^*$ on $xs$ and
$ys$ by aligning elements of $xs$ maximally with elements of $ys$ and then using $\delta_V$
on these pairs to produce a list $ds$ of numbers. In this process, some elements of
$ys$ will not be aligned if $length(xs) < length(ys)$, in which case we simply collect
these in another list of elements of $V$. For instance,
$$\delta_V^*([``Lovelace", ``Kleene", ``Turing"], [``Turing", ``Lovelace"]) = ([0,0], [``Kleene"])$$
since we align \lstinline|"Lovelace"| to \lstinline|"Lovelace"| and 
\lstinline|"Turing"| to \lstinline|"Turing"| for a difference of 0 in each case, which leaves
\lstinline|"Kleene"| unaligned.

>>>>

\newcommand{\kw}[1]{\ensuremath{\mathsf{#1}}\xspace}
\newcommand{\get}{\ensuremath{\kw{get}}\xspace}
\newcommand{\pput}{\ensuremath{\kw{put}}\xspace}
\newcommand{\create}{\ensuremath{\kw{create}}\xspace}
\newcommand{\res}{\ensuremath{\kw{res}}\xspace}
\newcommand{\ccopy}{\ensuremath{\kw{copy}}\xspace}
\newcommand{\const}{\ensuremath{\kw{const}}\xspace}
\newcommand{\default}{\ensuremath{\kw{default}}\xspace}
\newcommand{\length}{\ensuremath{\kw{length}}\xspace}
\newcommand{\match}{\ensuremath{\kw{match}}\xspace}
\newcommand{\filter}{\ensuremath{\kw{filter}}\xspace}
\newcommand{\strfilter}{\ensuremath{\kw{str\_filter}}\xspace}
\newcommand{\strunfilter}{\ensuremath{\kw{str\_unfilter}}\xspace}
\newcommand{\mergelens}{\ensuremath{\kw{merge}}\xspace}

\DeclareMathOperator*{\argmin}{argmin}
\DeclareMathOperator*{\argmax}{argmax}

@@

For example, assume that a source consists of a list of pairs of names, and that the target consists of a list of single names. Assume that $\ell$ is a bx which in the forward direction (i.e. the \get direction) simply discards the first of the two names:
$$\ell.\get \; [``Ada \; Lovelace", \; ``Alan \; Turing"] = [``Lovelace", \; ``Turing"]$$
If the order of the names in the target is reversed then a user might expect that in the backward direction (i.e. the \pput direction), the lens would simply restore the original names:
$$\ell.\pput \; [``Ada \; Lovelace", \; ``Alan \; Turing"] \; [``Turing", \; ``Lovelace"] = [``Ada \; Lovelace", \; ``Alan \; Turing"]$$
However, $\ell$ may not perform this alignment so that instead, the following result occurs:
$$\ell.\pput \; [``Ada \; Lovelace", \; ``Alan \; Turing"] \; [``Turing", \; ``Lovelace"] = [``Ada \; Turing", \; ``Alan \; Lovelace"]$$
Consider another example where both a source and a target consists of a list where every element on the list is a name and a year. Assume that $\ell$ is a bx which in the \get direction returns all elements of the source list that have year less than 1900:
\begin{align*}
&\ell.\get \; [``Ada \; Lovelace, \; 1915", \; ``Alan \; Turing, \; 1912", \; ``George \; Boole, \; 1815"] = [``George \; Boole, \; 1815"]
\end{align*}
Suppose that the entry $``Ada \; Lovelace, \; 1815"$ is then added to the target list. Should $\ell.\pput$ replace the old entry containing the name $``Ada \; Lovelace"$,
\begin{align*}
&\ell.\pput \; [``Ada \; Lovelace, \; 1915", \; ``Alan \; Turing, \; 1912", \; ``George \; Boole, \; 1815"]\\
&[``Alan \; Turing, \; 1912", \; ``Ada \; Lovelace, \; 1815"] \\
&= [``Ada \; Lovelace, \; 1815", \; ``Alan \; Turing, \; 1912", \; ``George \; Boole, \; 1815"]
\end{align*}
or should the \pput application retain both versions of $``Ada \; Lovelace"$?
\begin{align*}
&\ell.\pput \; [``Ada \; Lovelace, \; 1915", \; ``Alan \; Turing, \; 1912", \; ``George \; Boole, \; 1815"] \\
& [``Alan \; Turing, \; 1912", \; ``Ada \; Lovelace, \; 1815"]\\
&= [``Ada \; Lovelace, \; 1915", \; ``Alan \; Turing, \; 1912", \; ``George \; Boole, \; 1815", \; ``Ada \; Lovelace, \; 1815"]
\end{align*}
In both of these examples, there are multiple ways to restore consistency---that is, there are multiple ways to define $\ell.\pput \; s \; v$ in such a way that $\ell.\get \; (\ell.\pput \; s \; v) = v$---and the results from the different \pput applications may be vastly different from what a user expects.

@@

Each one of these solutions has significant shortcomings. While matching lenses successfully address the issue of alignment, they do not address the general issue of correct update propagation. Edit lenses and delta lenses all assume that the lens propagates a specific set of updates explicitly, but this is unrealistic since since in general users do not update data using a small predetermined langauge of edits. Instead users edit the data in any way, for example using text editors or IDEs, and present the updatad data to the bx wholesale. Finally, least change lenses are in general computationally intractable \cite{buneman2002propagation} and also lack a notion of composition---that is, the composition of two deterministic least change lenses is not always a least change lens\cite{macedo2013composing}.

We take the approach that instead of having a bx manually propagate a fixed set of edits, a bidirectional programming language should have explicit guarantees on how programs written in that language propagate updates; ideally these guarantees should be strictly stronger than hippocraticness. Consider again the first example that we presented, which is an instance of the alignment problem. In that example, the list
$[``Lovelace", \; ``Turing"]$ was updated to the list $[``Turing", \; ``Lovelace"]$. The programmer may wish to consider these two lists equivalent or not: if she does, then she should use a min-edit distance ``metric'' and demand that the source side respects this metric. On the other hand, if she wishes to use a positional metric to compare these two lists, then she should similarly demand that the source side respects this metric. In each case, the programmer does not care how the bx restores consistency; she cares only that the bx restores consistency and respects the appropriate metric.

In this paper, we present a general approach for describing update propapagtion guarantees for a bx without requiring that the bx manually handle updates. In particular, we will enhance a standard definition of bx transformations that only takes into account a consistency relation with a ``difference'' relation that describes a relationship between source updates and target updates. With this new formalization, a well behaved bx will not only be hippocratic, but will also be required to preserve the difference relation that is part of the bx.

We will then use this formalization to describe a particular language of bx transformations which we call {\em difference lenses}. Difference lenses are based on the {\em basic lenses} defined in the lens programming language {\em Boomerang} \cite{foster2009bidirectional}, but come with the guarantee that the lens respects ``differences'' on the source and view side of a lens. Intuitively, this means that if one starts with a source, maps it into a view, edits the view and then maps the updated view back into the old source, then the difference between the new and old source is at most the difference between the old and new view. Difference lenses will have combinators that deal with local alignment; in particular, difference lenses have positional, non-crossing, and min-edit distance iteration combinators that preserve positional, non-crossing, and min-edit distance differences respectively. We will then build on difference lenses to describe a language of {\em matching difference lenses} that deal with global alignment.

\sam{TODO: Talk about relational lenses.}

The major technical contributions that we will make in this paper are the following:
\begin{enumerate}
\item
We present a formalization for bx transformations that requires a bx to provide guarantees on how it propagates updates. 
\item
We describe a specific bx language which we call {\em difference lenses}; this language is based on the {\em basic lenses} defined in the lens programming language {\em Boomerang}\cite{foster2009bidirectional}, but each difference lens comes with explicit guarantees on update propagation. As a result of this work, we will describe non-trivial update propagation properties that Boomerang's basic lenses provide.
  \item
  We show how to represent matching lenses and  relational lenses \cite{bohannon2006relational} as difference lenses, and hence describe non-trivial guarantees that these lens languages provide as well.
\end{enumerate}
\section{Laying Down BX Laws}
\begin{definition}[Stevens \cite{stevens2010bidirectional}]
A bidirectional transformation (bx) $R : M \Leftrightarrow N$ comprises of 
\begin{enumerate}
\item
a consistency relation $R \subseteq M \times N$,
\item
restorers $\overrightarrow{R} : M \times N \rightarrow N$ and $\overleftarrow{R} : M \times N \longrightarrow
N$.
\end{enumerate}
\end{definition}

\begin{definition}
A bx $R : M \Leftrightarrow N$ is hippocratic if for all $m \in M$, $n \in N$, if $R(m, n)$ then $\overrightarrow{R}(m, n) = n$ and $\overleftarrow{R}(m,n) = m$. $R$ is {\em correct} if the consistency restorers do restore consistency: that is, for all $m \in M, n \in N$, we have $R(m, \overrightarrow{R}(m, n))$, and dually. $R$ is {\em well behaved} if it is correct and hippocratic.
\end{definition}
As we saw in the examples in the introduction, consistency restoration often times is too loose of a constrain to prevent a well-behaved bx from causing unexpected behaviour. What is missing from this model is how the bx handles non-trivial updates. The following definition factors in this missing piece, by relating updates on one side of a bx to updates on the other side of the bx.
\begin{definition}
A difference bidirectional transformation (diff-bx) $(R,S) : M \Leftrightarrow N$ comprises of 
\begin{enumerate}
\item
a consistency relation $R \subseteq M \times N$,
\item
a difference relation $S \subseteq (M \times M) \times (N \times N)$ such that for all $m, m' \in M$, $n, n' \in N$,
\begin{enumerate}
\item
if $(m, m') \stackrel{S}{\sim} (n, n)$ then $m = m'$,
\item
if $(m, m) \stackrel{S}{\sim} (n, n')$ then $n = n'$, and
\end{enumerate}
\item
restorers $\overrightarrow{R} : M \times N \rightarrow N$ and $\overleftarrow{R} : M \times N \longrightarrow
N$.
\end{enumerate}
\end{definition}
Intuitively, the relation $S$ relates differences on the $M$ side to differences on the $N$ side. The relation is required to send only trivial differences to trivial differences.
\begin{definition}
A diff-bx $(R,S) : M \Leftrightarrow N$ is {\em $S$-preserving} if for all $m \in M$, $n, n' \in N$, then if $R(m, n)$ then $(m, \overleftarrow{R}(m,n')) \stackrel{S}{\sim} (n, n')$
\end{definition}
\begin{fact}
If $(R,S) : M \Leftrightarrow N$ is $S$-preserving, then $R$ is hippocratic.
\end{fact}
\begin{definition}
A diff-bx $(R,S) : M \Leftrightarrow N$ is {\em well behaved} if $(R,S)$ is $S$-preserving and $R$ is correct.
\end{definition}
In the next section we describe a language for deriving well-behaved difference bidirectional transformations based on the {\em basic lenses} implemented in the lens programming language {\em Boomerang}.
\section{Difference Lenses}
\begin{definition}
A lens $\ell : S \Leftrightarrow V$ is a well behaved bx where the forward transformation $\overrightarrow{\ell} : S \times V \longrightarrow V$ does not depend on its $V$-argument, and where the consistency relation $\ell$ is given by $(s, v) \in R$ if and only if $\overrightarrow{\ell}(s) = v$

Equivalently, a lens $\ell : S \Leftrightarrow V$ is a pair of functions $\ell.\get : S \longrightarrow V$ and $\ell.\pput : S \longrightarrow V \longrightarrow V$ such that the following equations hold:
\begin{align*}
\ell.\get \; (\ell.\pput \; s \; v) &= v \tag{PUTGET}\\
\ell.\pput \; s \; (\ell.\get \; s) &= s \tag{GETPUT}
\end{align*}
\end{definition}
In order to strengthen hippocraticness (i.e. the GETPUT) law, we need to identify a difference relation $S$ that relates source updates to view updates. Our choice is relate the pair $(\ell.\pput \; s \; v, s)$ to the pair $(v, \ell.\get \; s)$ if and only if $\delta_S(\ell.\pput \; s \; v, s) \geq \delta_V(v, \ell.\get \; s)$ for some real-valued functions $\delta_S : S \times S \longrightarrow \mathbb{R}^{\geq 0}$ and $\delta_V : V \times V \longrightarrow \mathbb{R}^{\geq 0}$.

The functions $\delta_S$ and $\delta_V$ will depend only on $S$ and $V$, and will be designed to measure how similar two sources (or views) are to each other. More specifically, given a regular expression $R$ and strings $s, s' \in \mathcal{L}(R)$, we will define a number $\delta_R(s, s')$ inductively on $R$, with $0 \leq \delta_R(s, s') \leq 1$, and a high value of $\delta_R(s, s')$ indicates that $s$ and $s'$ are similar---in particular, if $s = s'$, then $\delta_R(s, s') = 1$, and if $s \neq s'$ and either $s = \varepsilon$ or $s' = \varepsilon$, then $\delta_R(s, s') = 0$. We will also expand our definition of the regular expression $R^*$, having three different variants $R^*_{POS}$, $R^*_{NC}$ and $R^*_{MED}$, where the differences are computed using positional, non-crossing and min-edit distance alignment strategies respectively, so that for all $s, s' \in \mathcal{L}(R^*)$, $\delta_{R^*_{MED}}(s, s') \geq \delta_{R^*_{NC}}(s, s') \geq \delta_{R^*_{MED}}(s, s')$.
\subsection{Difference Functions on Regular Languages}
\begin{definition}
Given a finite alphabet $\Sigma$, a {\em difference regular expression} is any expression generated by the following grammar,
$$R := s \; | \; R_1 \cdot R_2 \; | \; R_1 \cup R_2 \; | \; R^*_{POS} \; | \; R^*_{NC} \; | \; R^*_{MED}$$
where $s$ ranges over $\Sigma^*$ and $R$ ranges over regular expressions. The languages of $R^*_{POS}$, $R^*_{NC}$ and $R^*_{MED}$ are all the same, namely $\mathcal{L}(R^*)$.
\end{definition}
\begin{definition}
Given an unambiguous difference regular expression $R$, define $\delta_R : \mathcal{L}(R) \times \mathcal{L}(R) \longrightarrow \mathbb{R}^{\geq 0}$ inductively as follows:
\begin{align*}
\delta_s(t, t') &= 1\\
\delta_{R_1 \cdot R_2}(s_1 \cdot s_2, s'_1 \cdot s'_2) &= \frac{1}{2}(\delta_{R_1}(s_1, s'_1) + \delta_{R_2}(s_2, s'_2))\\
\delta_{R_1 \cup R_2}(s, s') &= 
\begin{cases}
\delta_{R_1}(s, s') & \text{if }s, s' \in \mathcal{L}(R_1) \\
\delta_{R_2}(s, s') & \text{if }s, s' \in \mathcal{L}(R_2)\\
0 & \text{otherwise }
\end{cases}\\
\delta_{R^*_{POS}}(s_1 \cdot \ldots \cdot s_m, t_1 \cdot \ldots \cdot t_n) &= 
\begin{cases}
0 & \text{if } m=0 \text{ or } n=0\\
\max\limits_{\{\varphi : [m] \longrightarrow [n] \; | \; \varphi(i) = i\}}\frac{1}{n}\sum\limits_{i=1}^{m}\ \delta_R(s_i, t_{\varphi(i)})& \text{if } m \leq n\\
\max\limits_{\{\varphi : [n] \longrightarrow [m] \; | \; \varphi(i) = i\}}\frac{1}{m}\sum\limits_{i=1}^{m}\ \delta_R(t_i, s_{\varphi(i)})& \text{otherwise}
\end{cases}\\
\delta_{R^*_{NC}}(s_1 \cdot \ldots \cdot s_m, t_1 \cdot \ldots \cdot t_n) &= 
\begin{cases}
0 & \text{if } m=0 \text{ or } n=0\\
\max\limits_{\{\varphi : [m] \longrightarrow [n] \; | \; \varphi \text{ is strictly increasing}\}}\frac{1}{n}\sum\limits_{i=1}^{m}\ \delta_R(s_i, t_{\varphi(i)})& \text{if } m \leq n\\
\max\limits_{\{\varphi : [n] \longrightarrow [m] \; | \; \varphi \text{ is strictly increasing}\}}\frac{1}{m}\sum\limits_{i=1}^{m}\ \delta_R(t_i, s_{\varphi(i)})& \text{otherwise}
\end{cases}\\
\delta_{R^*_{MED}}(s_1 \cdot \ldots \cdot s_m, t_1 \cdot \ldots \cdot t_n) &= 
\begin{cases}
0 & \text{if } m=0 \text{ or } n=0\\
\max\limits_{\{\varphi : [m] \longrightarrow [n] \; | \; \varphi \text{ is injective}\}}\frac{1}{n}\sum\limits_{i=1}^{m}\ \delta_R(s_i, t_{\varphi(i)})& \text{if } m \leq n\\
\max\limits_{\{\varphi : [n] \longrightarrow [m] \; | \; \varphi \text{ is injective}\}}\frac{1}{m}\sum\limits_{i=1}^{m}\ \delta_R(t_i, s_{\varphi(i)})& \text{otherwise}
\end{cases}
\end{align*}
\end{definition}
\begin{fact}
For any regular expression $R$ and $s, s' \in \mathcal{L}(R)$
\begin{enumerate}
\item
$0 \leq \delta_R(s, s') \leq 1$.
\item
$\delta_R(s, s) = 1$.
\item
If $s \neq s'$ and either $s = \varepsilon$ or $s' = \varepsilon$, then $\delta_R(s, s') = 0$
\item
For all $s, s' \in \mathcal{L}(R^*)$, $\delta_{R^*_{MED}}(s, s') \geq \delta_{R^*_{NC}}(s, s') \geq \delta_{R^*_{MED}}(s, s')$.
\item
$\delta_R(s, s') = \delta_R(s', s)$.
\end{enumerate}

\end{fact}
\subsection{Lens Combinators}
\begin{definition}
A lens $\ell : S \Leftrightarrow V$ is a {\em difference lens} if it satisfies the following equation:
\begin{equation}\label{difflenslaw}
\delta_S(\ell.\pput \; s \;v, s) \geq \delta_V(v, \ell.\get \; )
\end{equation}
\end{definition}
\cref{difflenslaw} can be expressed by a relation $S_{\ell} \subseteq (S \times S) \times (V \times V)$ by $(\ell.\pput \; s \;v, s)$ is related to $(v, \ell.\get \; s)$ if and only if \cref{difflenslaw} holds.

In the following part of the paper we describe a language for defining difference lenses on regular languages.
\subsubsection{Copy}
Let $R$ be any unambiguous difference regular expression. Define the lens $\ccopy(R) : R \Leftrightarrow R$ by
\begin{align}
\ccopy(R).\get \; s &= s\\
\ccopy(R).\pput \; s \; v &= v\\
\ccopy(R).\create \; v &= v
\end{align}
\begin{claim}
If $R$ is an unambiguous difference regular expression, then $\ccopy(R) : R \Leftrightarrow R$ is a difference lens.
\end{claim}
\begin{proof}
We need to show that $\delta_R(\ccopy(S).\pput
\; s \; v, s)) \geq \delta_R(v, \ccopy(S).\get \; s)$ which follows from the fact that $\delta_R(\ccopy(S).\pput \; s
\; v, s)) = \delta_R(v, s) = \delta_R(v, \ccopy(S).\get \; s) $.
\end{proof}
\subsubsection{Constant}
Let $R$ be any unambiguous difference regular expression and $c \in \Sigma^*$. Assume that $s' \in \mathcal{L}(R)$. Define the \const lens $\const(R, c) : R
\Leftrightarrow c$ by
\begin{align*}
\const.\get \; s &= c\\
\const.\pput \; s \; v &= s\\
\const.\create \; v &= s'
\end{align*}
\begin{claim}
If $R$ is an unabmbiguous difference regular expression, then $\const(R, c)$ is a difference lens.
\end{claim}
\begin{proof}
We need to show that $\delta_R(\const(R, c).\pput
\; s \; v, s)) \geq \delta_c(v, \const(R, c).\get \; s)$ which follows from the fact that $\delta_R(\const(R, c).\pput
\; s \; v, s)) = \delta_R(s, s) = 1 = \delta_c(c, c) = \delta_R(v, \const(R, c).\get \; s)$.
\end{proof}

\subsubsection{Default}
Let $\ell : S \Leftrightarrow V$ be a difference lens. Let $f :
V \longrightarrow S$ be any function. Define the lens $\default(\ell, f) :
S \Leftrightarrow V$ by
\begin{align*}
\default(\ell, f).\get \; s &= \ell.\get \; s\\
\default(\ell, f).\pput \; s \; v &= \ell.\pput \; s \; v\\
\default(\ell, f).\create \; v &= \ell.\pput \; (f \; v) \; v
\end{align*}
